Squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-24 large_number_squeezing_with_atoms.tex
Go thunderbirds, go!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
  \expandafter\let\csname equation*\endcsname\relax
  \expandafter\let\csname endequation*\endcsname\relax
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{amsmath}
% \usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\captionsetup{compatibility=false}
\usepackage{subcaption}

@@

\title{Large squeezing in Bose-Einstein condensates with large numbers of atoms}

@@

The creation, detection and application of squeezing requires precise mode-matching, and is therefore very sensitive to uncontrolled coupling between spatial modes of the atomic field.  Atomic squeezing experiments have typically attempted to operate in the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}.  However, tight traps limit the total number of atoms, as the 3-body recombination rate places an effective maximum atomic density in the gas, and well before that limit the interatomic interactions break the single-mode operation.  Both of these effects limit the number of atoms that can be effectively squeezed, so we will examine a broader range of trap geometries, and calculate the dynamics in the presence of multiple spatial modes. Many similar previous simulations have been performed in one or two spatial dimensions, by integrating out some of the remaining dimensions \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, kheruntsyanET2002}.  For simulations of semiclassical fields this can be a reasonable approximation, but it can ignore spontaneous scattering processes in full quantum field theory calculations, so we take care to include full 3D simulations.

In Section \ref{secTwoModeAnalytic} we examine the ideal behaviour of a nonlinear bosonic two-state system when that system possesses only a single spatial mode, extracting full analytic solutions for the number squeezing in each of the two states, as well as the number difference squeezing between the two states. These solutions of the ideal case provide a check on the multimode cases we consider later in the paper, as well as providing a bound on the best possible squeezing for the system.

In Section \ref{sec:MMdescription} we show how multimode behaviour becomes important in physical systems, and introduce our model for simulating higher dimensional trap geometries.  Section~\ref{sec:1Dsim} analyses one-dimensional models, highlighting some of the issues regarding optimising squeezing parameters and mode-matching in squeezing experiments.  Section~\ref{sec:pipulse} shows that the application of a $\pi$-pulse during the process can significantly improve mode-matching issues.  Section~\ref{sec:2D3Dsqueezing} investigates measurable squeezing in 2D and 3D, and uses a Bogoliubov analysis to explain the scaling with system size and dimensionality. Section~\ref{sectionConclusion} describes the conclusion that large volume, flat-bottomed, three-dimensional traps, which are obviously compatible with large atom number, can produce extremely high degrees of squeezing.

@@

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/zero_effective_chi.eps}
    \caption{Relative number squeezing in state $|a\rangle$ (solid line) and number difference squeezing between states $|a\rangle$ and $|b\rangle$ (dashed line) as a function of final recombination time in the case where $\chi_+ = \chi_{aa} + \chi_{bb} - 2 \chi_{ab} = 0$. Parameters: $n_a = n_b = 2 \times 10^5$, $\chi_{aa}=0.03\, \text{s}^{-1}, \chi_{ab}=0.02, \chi_{bb}=0.01\,\text{s}^{-1}, \tau_{\mathrm{hold}}=2\times 10^{-3}$s. Solid line, $\phi=1.67$; dashed line, $\phi=1.55$.} 
    \label{figZeroEffectiveChi}
\end{figure}

@@

Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to apply a $\pi$-pulse to swap the populations of the internal states after half of the evolution.  As can be seen by Eqs.~\eqref{eqa1} and \eqref{eqb1}, such a pulse also leaves the quantum statistics of each state unchanged.  The idea of this scheme is to make the effective potential seen by the atoms in state $\ket{a}$ during the second half of the hold time very similar to that seen by the atoms that were in state $\ket{a}$ during the first half of the hold time, and will only work if the modes $\ket{a}$ and $\ket{b}$ have an equal occupation at the start of the hold time, or if $\chi_+=\chi_{aa} + \chi_{bb} - 2\chi_{ab}=0$.  In the latter case, this reduces to a spin-echo pulse, and as the squeezing effects of the $\Delta\omega$ term in Eq.~\eqref{eqNonlinearJosephson} are cancelled in this case, there will be no squeezing.  However, in the limit of short hold times, for equal occupations of the two modes, and when $\chi_+\neq 0$, this technique reduces the difference in the mode shapes after the hold time.  In practice this will never work perfectly --- due to the density distributions of $\psi_a(\mathbf{r})$ and $\psi_b(\mathbf{r})$ not being in the ground state, their mode shapes will have changed by the time the $\pi$-pulse is applied, and so the spatial dynamics in the second half of the hold time will not be quite the same as those during the first half.  Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement.  More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
  \expandafter\let\csname equation*\endcsname\relax
  \expandafter\let\csname endequation*\endcsname\relax
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{cite}
% \usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\captionsetup{compatibility=false}
\usepackage{subcaption}

@@

\title{Squeezing in Bose-Einstein condensates with large numbers of atoms}

@@

The creation, detection and application of squeezing requires precise mode-matching, and is therefore very sensitive to uncontrolled coupling between spatial modes of the atomic field.  Atomic squeezing experiments have typically attempted to operate in the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}.  However, tight traps limit the total number of atoms, as the 3-body recombination rate places an effective maximum atomic density in the gas, and well before that limit, the interatomic interactions break the single-mode operation.  Both of these effects limit the number of atoms that can be squeezed effectively, so we will examine a broader range of trap geometries, and calculate the dynamics in the presence of multiple spatial modes. Many similar previous simulations have been performed in one or two spatial dimensions, by integrating out some of the remaining dimensions \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, kheruntsyanET2002}.  For simulations of semiclassical fields this can be a reasonable approximation, but it can ignore spontaneous scattering processes in full quantum field theory calculations, so we take care to include full 3D simulations.

In Section \ref{secTwoModeAnalytic} we examine the ideal behaviour of a nonlinear bosonic two-state system when that system possesses only a single spatial mode, extracting full analytic solutions for the number squeezing in each of the two states, as well as the number difference squeezing between the two states. These solutions of the ideal case provide a check on the multimode cases we consider later in the paper, as well as providing a bound on the best possible squeezing for the system.  In Section \ref{sec:MMdescription} we show how multimode behaviour becomes important in physical systems, and introduce our model for simulating higher dimensional trap geometries.  Section~\ref{sec:1Dsim} analyses one-dimensional models, highlighting some of the issues regarding optimising squeezing parameters and mode-matching in squeezing experiments.  Section~\ref{sec:pipulse} shows that the application of a $\pi$-pulse during the process can significantly improve mode-matching issues.  Section~\ref{sec:2D3Dsqueezing} investigates measurable squeezing in 2D and 3D, and uses a Bogoliubov analysis to explain the scaling with system size and dimensionality. Section~\ref{sectionConclusion} describes the conclusion that large volume, flat-bottomed, three-dimensional traps, which are obviously compatible with large atom number, can produce extremely high degrees of squeezing.

@@

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/zero_effective_chi.eps}
    \caption{Number squeezing in state $|a\rangle$ (solid line) and number difference squeezing between states $|a\rangle$ and $|b\rangle$ (dashed line) as a function of final recombination time in the case where $\chi_+ = \chi_{aa} + \chi_{bb} - 2 \chi_{ab} = 0$. Parameters: $n_a = n_b = 2 \times 10^5$, $\chi_{aa}=0.03\, \text{s}^{-1}, \chi_{ab}=0.02, \chi_{bb}=0.01\,\text{s}^{-1}, \tau_{\mathrm{hold}}=2\times 10^{-3}$s. Solid line, $\phi=1.67$; dashed line, $\phi=1.55$.} 
    \label{figZeroEffectiveChi}
\end{figure}

@@

Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to apply a $\pi$-pulse to swap the populations of the internal states after half of the evolution.  As can be seen by Eqs.~\eqref{eqa1} and \eqref{eqb1}, such a pulse also leaves the quantum statistics of each state unchanged.  The idea of this scheme is to make the effective potential seen by the atoms in state $\ket{a}$ during the second half of the hold time very similar to that seen by the atoms that were in state $\ket{a}$ during the first half of the hold time, and will only work if the modes $\ket{a}$ and $\ket{b}$ have an equal occupation at the start of the hold time, or if $\chi_+=\chi_{aa} + \chi_{bb} - 2\chi_{ab}=0$.  In the latter case, this reduces to a spin-echo pulse, and as the squeezing effects of the $\Delta\omega$ term in Eq.~\eqref{eqNonlinearJosephson} are cancelled in this case, there will be no squeezing.  However, in the limit of short hold times, for equal occupations of the two modes, and when $\chi_+\neq 0$, this technique reduces the difference in the mode shapes after the hold time.  In practice this will never work perfectly, as $\psi_a(\mathbf{r})$ and $\psi_b(\mathbf{r})$ will typically have evolved by the time the $\pi$-pulse is applied, and so the spatial dynamics in the second half of the hold time will not be quite the same as those during the first half.  Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement.  More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.
